---
title: "Using tntpr data cleaning functions"
date: '`r Sys.Date()`'
output:
  github_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,echo=FALSE, include = FALSE}
library(knitr)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(out.width='750px', dpi = 300)
knitr::opts_chunk$set(dev = "png", fig.width = 8, fig.height = 4.8889, dpi = 300)
knitr::opts_chunk$set(fig.path = "introduction_files/")
```

```{r load_packages, include = FALSE}
library(pacman)
if(!require("tntpr")) install_github("tntp/tntpr"); library(tntpr)
p_load(tidyverse, janitor)
```

Common Scenerio

Imagine receiving the `wisc_dirtied` dataset from a district. While the actual `wisc` dataset has nicely formatted data types, this version does not. Notice that all the the variables are either numeric or characters.

```{r, echo=FALSE, include=FALSE}
wisc_dirtied <- wisc %>% 
  select(student_id, grade, district, school, year, attday, readSS, mathSS, proflvl, race) %>% 
  mutate_at(vars(grade, district, school, year, proflvl, race), as.character)
```

```{r}
wisc_dirtied %>% 
  glimpse()
```

One of the first steps in the data cleaning process is to "factorize" the dataset. R uses factors to handle categorical variables, variables that have a fixed and known set of possible values. Usually this one of the most manual, error-prone parts of the data-cleaning process.

Below is a workflow I use that might be helpful for others. I use a couple functions that may be new to you.


- `factorize_df`: examines each column in a data.frame; when it finds a column composed solely of the values provided to the \code{lvls} argument it updates them to be factor variables, with levels in the order provided.



Step 1: Identify which variables are factors.This part is 

```{r}

wisc_all_character_data_types %>% 
  map(unique) %>% 
  map(length)

```



