---
title: 'Worked Example: `tntpr::date_to_sy`'
output:
  html_document:
    keep_md: yes
    toc: no
  word_document:
    toc: no
vignette: |
  %\VignetteIndexEntry{Vignette Title} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r setup,echo=FALSE, include = FALSE}
library(knitr)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(out.width='750px', dpi = 300)
knitr::opts_chunk$set(dev = "png", fig.width = 8, fig.height = 4.8889, dpi = 300)
```

```{r load_packages, include = FALSE}
library(pacman)
if(!require("tntpr")) install_github("tntp/tntpr"); library(tntpr)
p_load(tidyverse, janitor, lubridate)
```

- `date_to_sy`: Checks to see if a date is past the user-specified cutoff point for delineating school years, then maps to the appropriate year.

Say you have a test date and you want to create a new variable that tells you the school year the test was taken.


```{r, echo=FALSE}
set.seed(1)

appl_dat <- tibble(student_id = 1:100, 
                   test_date = sample(seq(as.Date('2010/08/01'), as.Date('2020/01/01'), by="day"), 100))

appl_dat

```

Historically, I would have used a long, error-prone `case_when` mutation, but the `tntpr::date_to_sy` function is much easier.

The function takes two arguments, `date_var` and `last_day_of_sy` (year doesn't matter) and returns a character string with the school year in the form '(year) - (year)'. 

```{r}
appl_dat %>% 
  mutate(hire_date_sy = date_to_sy(test_date, last_day_of_sy = ymd("2018-06-01")))
```
