---
title: "Factorizing a survey dataset"
date: '`r Sys.Date()`'
output:
  github_document:
    toc: false
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,echo=FALSE, include = FALSE}
library(knitr)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(out.width='750px', dpi = 300)
knitr::opts_chunk$set(dev = "png", fig.width = 8, fig.height = 4.8889, dpi = 300)
knitr::opts_chunk$set(fig.path = "introduction_files/")
```

```{r load_packages, include = FALSE}
library(pacman)
if(!require("tntpr")) install_github("tntp/tntpr"); library(tntpr)
p_load(tidyverse, janitor, praise)
```

```{r example_data, echo=FALSE, include=FALSE}
qtype_1 <- c("Strongly Disagree", "Disagree", "Somewhat disagree", "Somewhat agree", "Agree", "Strongly agree")
qtype_2 <- c("Strongly Disagree", "Disagree", "Neither agree nor disagree", "Agree", "Strongly agree")
qtype_3 <- c("No", "Yes")

survey_dat <- tibble(response_id = 1:100, 
                     years_of_experience = round(runif(100) * 10, digits = 0), 
                     q1 = sample(qtype_1, 100, replace = TRUE), 
                     q2 = sample(qtype_1 %>% str_remove("Strongly disagree"), 100, replace = TRUE), 
                     q3 = sample(qtype_1, 100, replace = TRUE), 
                     q4 = replicate(100, praise()), 
                     q5 = sample(qtype_2, 100, replace = TRUE), 
                     q6 = sample(qtype_2, 100, replace = TRUE), 
                     q7 = replicate(100, praise("${Exclamation}! ${EXCLAMATION}!-${EXCLAMATION}! This is just ${adjective}!")),
                     q8 = sample(qtype_3, 100, replace = TRUE),
                     q9 = sample(qtype_3, 100, replace = TRUE))

```

## Worked Example

Imagine you conduct a survey and are about to analyze the data. One of the first steps in the data cleaning process is to "factorize" the dataset. R uses factors to handle categorical variables, variables that have a fixed and known set of possible values. Usually this is one of the most manual, error-prone parts of the data-cleaning process.

Below is a workflow I use that might be helpful for others. I use a couple functions that may be new to you.

- `glimpse`: This function makes it possible to see every column in a data frame. It shows columns run down the page and data runs across.

- `map`: The purrr version of the apply functions. Transforms the input by applying a function to each element and returning a vector the same length as the input. In this example, I use the map function with a data frame as an arguement. In this case, the inputs are the variable vectors and map applies the function across these variable vectors.

- `factorize_df`: examines each column in a data.frame; when it finds a column composed solely of the values provided to the `lvls` argument it updates them to be factor variables, with levels in the order provided.

### Step 1: Identify which variables should be factors.

Admittedly, this part usually takes some knowledge of the dataset and/or exploratory perusing of the dataset. Let's first use `glimpse` to take a look at the dataset.

```{r}
survey_dat %>% 
  glimpse()
```

First thing I notice is all the question variables are character vectors. Since I know some of the questions where multiple-choice questions, it's likely some of these should be converted to factors.

One thing you could try is seeing how many unique values each variable has. 

```{r}
survey_dat %>% 
  map(unique) %>% 
  map(length)
```

Here I notice `response_id`, `q4`, and `q7` have 50+ unique values and should likely not be transformed to factors. Knowing the dataset, I think `response_id` should probably be a character data type but will not do that in this example.  Finally, since each `q1`, `q2`, `q3`, `q5`, `q6`, `q8`, and `q9` are "variables that have a fixed and known set of possible values" they are likely candidates to be factors.

Next, I'll see what the unique values are for these variables.

```{r}
survey_dat %>% 
  select(q1, q2, q3, q5, q6, q8, q9) %>% 
  map(unique)
```

Aha! These do indeed look like factors.

### Step 2: Transform to factors with appropriate level ordering.

Here a lot of people would `mutate` their dataset with a series of factor(., levels = c(...)) mutations. This works, but let's use `factorize_df` to help us do this faster with less typing.

First, let's see what the "levels" of `q1` are.

```{r}
survey_dat$q1 %>% unique()
```

Okay, looks like a 6-pt likert agreement scale. Let's pass these levels through the `factorize_df` function (in order) as the lvls arguement.

```{r}
survey_dat <- survey_dat %>% 
  factorize_df(lvls = c("Strongly Disagree", "Disagree", "Somewhat disagree", "Somewhat agree", "Agree", "Strongly agree"))

survey_dat %>% 
  glimpse()

survey_dat %>% 
  map(levels)
```

Okay, that doesn't look too impressive but notice that the function: 
  a) searched each column in the dataset and tested whether the unique values that were a subset of the values in the lvls arguement, 
  b) identified `q1`, `q2`, and `q3` as having fitting that criteria, and
  c) transformed those and only those three columns to factors with the appropriate levels

Next I can move onto the next factor `q5`.

```{r}
survey_dat$q5 %>% unique()
```

```{r}
survey_dat <- survey_dat %>% 
  factorize_df(lvls = c("Strongly Disagree", "Disagree", "Neither agree nor disagree", "Agree", "Strongly agree"))
```

And finally `q8`.

```{r}
survey_dat$q8 %>% unique()
```

```{r}
survey_dat <- survey_dat %>% 
  factorize_df(lvls = c("No", "Yes"))
```

### Step 3: Check if factorizing worked like you intended.

Okay we said we needed to change `q1`, `q2`, `q3`, `q5`, `q6`, `q8`, and `q9` to factors. Let's verify this worked.

```{r}
survey_dat %>% 
  select(q1, q2, q3, q5, q6, q8, q9) %>% 
  map(is.factor)
```

They're all factors...

```{r}
survey_dat %>% 
  select(q1, q2, q3, q5, q6, q8, q9) %>% 
  map(levels)
```

... with the correct levels and level order. And voila, a 'factorized' dataset.
