---
title: "Using the tntpr package"
date: "`r Sys.Date()`"
output:
  rmarkdown::github_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,echo=FALSE, include = FALSE}
knitr::opts_chunk$set(error = TRUE)
# knitr::opts_chunk$set(dev = "png", fig.width = 8, fig.height = 4.8889, dpi = 300)
```

## About

The `tntpr` package makes data science at TNTP easier and more accurate by supplying tools that are needed for common TNTP analyses.  Because this package serves just the TNTP analysis community, functions can be tailored very specifically to our exact use cases.

## Package summary

Some of the highlights of the package include:

- TNTP-themed RMarkdown templates, for starting a new analysis with a shell that can already generate a TNTP-themed .docx report 
- TNTP-specific ggplot2 theme and color palette
- Wrappers for quickly making typical TNTP-style charts (e.g., bar chart of the distribution of one variable, grouped by a second)
- Education-specific data management functions (e.g., `date_to_SY()` to convert continuous hire dates into school years using a specified cutoff date), and a built-in fake student achievement dataset to play with called `wisc`.

## Installing the package

This package is not on CRAN, and probably will not ever be.  You'll need to install this package from its Bitbucket repository.  You can add this to the top of your analysis script:

```
# install tntpr if you don't already have it, then load it
library(devtools) # for install_git()
if(!require("tntpr")) install_git("https://tools.tntp.org/bitbucket/scm/ct/tntpr.git")
library(tntpr)
```

This won't check for updates, though.  So for now, consider periodically running the `install_git(...)` part anyway.  Once it's installed, you can load it however you normally load packages, say with `pacman::p_load(tntpr)`.


## Usage

### Reporting templates

Start your analysis with a good-looking .docx file as output, right off the bat.  Right now we have just a single TNTP template, "Data Memo", but it can be adapted and improved and if we have other common needs (a different set of headings?) those can easily be separate templates.

**To access templates once you've installed the tntpr package:** go to `File` -> `New File` -> `R Markdown` -> `From Template`.  You'll see a choice "Data Memo" from the tntpr package.  Just specify the document's file name and the directory you want it in (ideally a Bitbucket repository) and you're off!

A file `tntp-style-file.docx` will be copied into that directory; leave it there.  That provides the TNTP .docx stylings when you re-knit your R Markdown document.

### Chart styles

```{r load_packages, include = FALSE}
library(pacman)
if(!require("tntpr")) install_git("https://tools.tntp.org/bitbucket/scm/ct/tntpr.git"); library(tntpr)
p_load(ggplot2, dplyr, janitor, knitr)
```

#### TNTP colors

You can access the official TNTP-branded colors using `palette_tntp()`.  This will return a vector with hex code for our colors:
```{r colors}
palette_tntp("dark_blue", "orange", "light_gray")
```

#### TNTP theme for ggplot2

Add `+ theme_tntp()` to your ggplot2 calls to format a chart in the common TNTP style.  For instance, this will use the Segoe UI font that is standard at TNTP.

```{r theme_tntp}
ggplot(mtcars, aes(x = cyl)) +
  geom_bar(fill = palette_tntp("dark_blue")) +
  theme_tntp()
```

### Chart building

#### Custom charting functions

So far there's the function `bar_chart_counts`, for making the ubiquitous TNTP-style bar chart of a distribution.  It automates some tedious aspects of chart-making that R is slow at, like text labels showing %s.  These charts are deck-ready, faster than Excel.

Here are examples with the built-in `wisc` data set (of fake Wisconsin test data):

```{r chart_functions}
bar_chart_counts(wisc, proflvl, var_label = "Proficiency Level")
bar_chart_counts(wisc,
                 race,
                 proflvl,
                 var_label = "Race",
                 labels = "pct",
                 title = "Distribution of Proficiency Levels by Student Race") +
  theme(plot.margin=unit(c(0,0,0,0),"cm"))
```

The function has lots of customization options, see `?bar_chart_continuous` for more.

We hope to add a similar function for showing means.

### Data management

The function `date_to_SY` cuts continuous dates, like hire dates, into discrete school years, at a specified cut-off date.  For instance, here are some hire dates, split into their school year hiring seasons based on a September 1st cut-off:
```{r}
hire_dates <- c(as.Date("2015-08-31"), as.Date("2015-10-20"), as.Date("2016-08-08"), as.Date("2016-08-31"), as.Date("2016-09-15"))
date_to_sy(hire_dates, last_day_of_sy = as.Date("2000-09-01")) # the year doesn't matter
```

We can incorporate other functions that have been written but not yet standardized for the package.  E.g., converting survey responses to top-2 agree:
```{r}
convert_to_top_2_agree <- function(x, custom_vals = NULL){
  if(is.null(custom_vals)){
    custom_vals <- c("strongly agree", "agree", "highly satisfied", "extremely satisfied", "satisfied", "very confident", "confident")
    x <- tolower(x)
  }
  if_else(is.na(x), as.character(NA),
          if_else(x %in% custom_vals, "Top-2 Agree", "Not in Top-2"))
}

convert_to_top_2_agree(c("Strongly agree", "Agree", "Somewhat agree", NA, NA, "Strongly disagree")) %>%
  tabyl
```

Or functions to discern teacher movement when comparison rosters and studying retention and transfer patterns.  This can be simple:

```{r}
# Write function to determine whether someone left the location the next year
left_campus <- function(first_year, second_year){
  if_else(is.na(first_year), NA,
         if_else(is.na(second_year), TRUE,
                if_else(first_year != second_year, TRUE, FALSE)))
}
```

Or complex:

```{r}
# From Noble St. contract, July 2016, Danielle Proulx
# Formula for detailing yearly movement
yearly_movement <- function(first_year, second_year){
  case_when(
    is.na(first_year) & is.na(second_year) ~ "Not Present",
    is.na(first_year) & second_year == "Teacher" ~ "Joined as Teacher",
    is.na(first_year) & second_year == "Leader" ~ "Joined as Leader",
    is.na(first_year) & second_year == "Other" ~ "Joined as Other"
    ,
    first_year == "Teacher" & is.na(second_year) ~ "Teacher Left",
    first_year == "Leader" & is.na(second_year) ~ "Leader Left",
    first_year == "Other" & is.na(second_year) ~ "Other Left",
    
    first_year == "Teacher" & second_year == "Teacher" ~ "Remained Teacher",
    first_year == "Teacher" & second_year == "Leader" ~ "Teacher became Leader",
    first_year == "Teacher" & second_year == "Other" ~ "Teacher became Other",
    
    first_year == "Leader" & second_year == "Leader" ~ "Remained Leader",
    first_year == "Leader" & second_year == "Teacher" ~ "Leader became Teacher",
    first_year == "Leader" & second_year == "Other" ~ "Leader became Other",
   
    first_year == "Other" & second_year == "Teacher" ~ "Other became Teacher",
    first_year == "Other" & second_year == "Leader" ~ "Other became Leader",
    first_year == "Other" & second_year == "Other" ~ "Remained Other",
    TRUE ~ "bad formula!!!"
  )
}
```