---
title: "tntp-style-plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tntp-style-plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(tntpr)

# Load packages - first pacman, installing if necessary, then others
if (!require("pacman")) install.packages("pacman")
library(pacman)
p_load(devtools, tidyverse, knitr)
if (!require("patchwork")) devtools::install_github("thomasp85/patchwork")
```

```{r knitr_options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(out.width = "750px", dpi = 300)
knitr::opts_chunk$set(dev = "png", fig.width = 8, fig.height = 4.8889, dpi = 300)
```

```{r sample_datasets, include=FALSE}
performance_data <- data.frame(
  teacher_experience = c(rep("0-3 years ", 5), rep("4-6 years", 4), rep("7+ years", 4)),
  y1_teacher_performance = sample(0:100, size = 13, rep = TRUE),
  y2_teacher_performance = sample(0:100, size = 13, rep = TRUE)
) %>%
  mutate(y1_performance_quartile = ntile(x = y1_teacher_performance, n = 4))

survey_question <- data.frame(
  question = c(rep("To what extent do you \nagree with ...", 100)),
  answer = sample(1:5, size = 100, rep = TRUE)
)
```


### ggplot Themes

tntpr now has two ggplot2 themes; the classic `theme_tntp()`, and the updated `theme_tntp_2018()`.

```{r theme_tntp, echo=FALSE, warning=FALSE}
ex_plot_default <- performance_data %>%
  ggplot(aes(factor(teacher_experience))) +
  geom_bar() +
  labs(
    title = "default",
    subtitle = "Subtitle",
    x = "x label",
    y = "y label",
    caption = "caption"
  )

ex_plot_theme_tntp <- performance_data %>%
  ggplot(aes(factor(teacher_experience))) +
  geom_bar() +
  labs(
    title = "theme_tntp()",
    subtitle = "Subtitle",
    x = "x label",
    y = "y label",
    caption = "caption"
  ) +
  theme_tntp()

ex_plot_theme_tntp_2018 <- performance_data %>%
  ggplot(aes(factor(teacher_experience))) +
  geom_bar() +
  labs(
    title = "theme_tntp_2018()",
    subtitle = "Subtitle",
    x = "x label",
    y = "y label",
    caption = "caption"
  ) +
  theme_tntp_2018()
```

```{r, echo=FALSE, warning=FALSE, fig.width=7}
ex_plot_default
```

```{r, echo=FALSE, warning=FALSE, fig.width=7}
ex_plot_theme_tntp
```

```{r, echo=FALSE, warning=FALSE, fig.width=7}
ex_plot_theme_tntp_2018
```

### palette_tntp() gives you access to TNTP-style colors:

```{r plot_color_palette, include = FALSE}
# This function is for displaying color palettes
plot_color_palette <- function(dat, title = "") {
  dat %>%
    as.data.frame() %>%
    rownames_to_column() %>%
    set_names(c("labels", "hex")) %>%
    mutate(row = row_number()) %>%
    ggplot(aes(x = row, y = 1, fill = hex)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = paste0(labels, " (", hex, ")")), position = position_stack(vjust = 0.5), color = "white") +
    scale_fill_identity() +
    coord_flip() +
    labs(
      title = title,
      x = NULL,
      y = NULL
    ) +
    theme_tntp_2018() +
    theme(
      axis.text = element_blank(),
      panel.grid = element_blank()
    )
}
```

You can still use palette_tntp the way you used to...

```{r}
palette_tntp("dark_blue")
```

... but now you have a larger selection of colors.

```{r, fig.height= 12, fig.width=5, warning=FALSE, echo = FALSE}
# A one-off function to plot color palettes, see code above if curious
plot_color_palette(tntpr::colors_tntp, title = "palette_tntp colors")
```

The `palette_tntp_scales()` function provides access to 5 TNTP color scales:

-- `"default"` (colors in the PPT and Word template), 

-- `"likert_4pt"`,`"likert_5pt"`, `"likert_6pt"`,and 

-- `"likert_orange_to_green_4pt"`, `"likert_orange_to_green_5pt"`, `"likert_orange_to_green_6pt"`

-- `colors_tntp_classic` (original `palette_tntp` colors from when this package was created years ago).


```{r, fig.height= 12, fig.width=7, warning=FALSE, echo = FALSE}
palettes <- c("default", "likert_4pt", "likert_5pt", "likert_6pt", "likert_orange_to_green_4pt", "likert_orange_to_green_5pt", "likert_orange_to_green_6pt")

palette_plots <- map(palettes, palette_tntp_scales) %>%
  map2(palettes, plot_color_palette)

palette_plots[[1]] + (palette_plots[[2]] + palette_plots[[3]] + palette_plots[[4]]) + palette_plots[[5]] + palette_plots[[6]] + palette_plots[[7]] + plot_layout(ncol = 1)
```

### `scale_fill_tntp()` and `scale_color_tntp()`

Supply TNTP-palette scales for filling and coloring.

```{r, fig.width=7, fig.length=5, warning=FALSE, echo=TRUE}
performance_data %>%
  ggplot(aes(factor(teacher_experience), fill = factor(y1_performance_quartile))) +
  geom_bar(position = position_fill()) +
  labs(
    title = "Title",
    subtitle = "Subtitle",
    x = "x label",
    y = "y label",
    fill = "fill",
    caption = "caption"
  ) +
  theme_tntp_2018() +
  scale_fill_tntp()
```

```{r, fig.width=7, fig.length=5, warning=FALSE, echo=TRUE}
performance_data %>%
  ggplot(aes(x = y1_teacher_performance, y = y2_teacher_performance, color = factor(teacher_experience))) +
  geom_point(size = 2) +
  labs(
    title = "Title",
    subtitle = "Subtitle",
    x = "x label",
    y = "y label",
    fill = "fill",
    caption = "caption"
  ) +
  theme_tntp_2018() +
  scale_color_tntp()
```


You can specify which color palette you want to use.
```{r, fig.width=7, fig.length=5, warning=FALSE, echo = TRUE}
survey_question %>%
  ggplot(aes(factor(question), fill = factor(answer))) +
  geom_bar(position = position_fill()) +
  labs(
    title = "Title",
    subtitle = "Subtitle",
    x = "x label",
    y = "y label",
    caption = "caption"
  ) +
  theme_tntp_2018() +
  scale_fill_tntp(palette = "likert_5pt")
```
