---
title: "tntpr-introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tntpr-introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(tntpr)
library(tidyverse)
```

## About

The `tntpr` package makes data science at TNTP easier and more accurate by 
supplying tools that are needed for common TNTP analyses.

## Package summary

Some of the highlights of the package include:

- A TNTP-themed RMarkdown template, for starting a new analysis with a shell that can already generate a TNTP-themed .docx report 
- Functions for initializing a new repository or project folder with TNTP-standard directories and documentation 
- TNTP-specific ggplot2 themes and color palettes
- Survey analysis tools
- Wrappers for quickly making typical TNTP-style charts (e.g., bar chart of the distribution of one variable, grouped by a second)
- Education-specific data management functions (e.g., `date_to_SY()` to convert continuous hire dates into school years using a specified cutoff date), and a built-in fake student achievement dataset to play with called `wisc`.

## Installing the package

This package is not on CRAN, and probably will not ever be.  You'll need to install this package from its GitHub repository.  You can add this to the top of your analysis script:

```{r, include = FALSE}
library(tntpr)
library(magrittr)
library(ggplot2)
library(tidyr)
library(dplyr)
```

Once installed, you can update the package with `update_tntpr()`.

## Usage

### Reporting templates

Start your analysis with a good-looking .docx file as output, right off the bat.
Right now we have just a single TNTP template, "Data Memo", but it can be 
adapted and improved and if we have other common needs (a different set of 
headings?) those can easily be separate templates.

**To access templates once you've installed the tntpr package:** go to `File` -> `New File` -> `R Markdown` -> `From Template`.  You'll see a choice "Data Memo" from the tntpr 
package.  Just specify the document's file name and the directory you want it 
in (probably a subfolder of a Bitbucket repository) and you're off!

A file `tntp-style-file.docx` will be copied into that directory; leave it there.
That provides the TNTP .docx stylings when you re-knit your R Markdown document.

### Setting up repositories and subfolders
This saves time getting started with a new analysis and encourages use of common 
file storage conventions and documentation that make work more browsable and transparent.

**Directory conventions**

TNTP's Bitbucket directory structure is a single repository per client or region, with 
subdirectories in the repository corresponding to specific analysis projects. For example:

> - <City #1>
>   - <City abbreviation> <Project #1 name>
>   - <City abbreviation> <Project #2 name>
>   - <City abbreviation> <Project #3 name>
> - <City or Region #2>
>   - ...

**Usage**

After creating a new, empty repo in Bitbucket and cloning it, run `setup_repo` to 
initialize it.  This will create a subfolder as well, in which you'll conduct an analysis project.

If the repository already exists, and you just want to begin a new analysis project, 
create the new subdirectory with `setup_subdirectory`.

Both functions take the same arguments, used to setup the project subfolder and its README:  
 -  *subfolder*: what the name of the subdirectory should be, e.g., "xyz_instructional_audit"  
 -  *proj_name*: the full name of the analysis project, e.g., "XYZ Public Schools Equity Study".  Appears in the README.  
 -  *analyst_name*: the analyst(s) working on this project.  Appears in the README.  

`setup_repo` will also add a `.Rproj` RProject file, `.gitignore` file, and create a README.Md file for the main repository.

### TNTP colors

You can access the official TNTP-branded colors using `palette_tntp()`.  This will 
return a vector with hex code for our colors:
```{r colors}
palette_tntp("dark_blue", "orange", "light_gray")
```

Or you can select a specific TNTP palette (`"default"`, `"colors_tntp_classic"`, `"likert_4pt"`, 
`"likert_5pt"`, or `"likert_6pt"`) with `palette_tntp_scales` and return a vector with hex 
codes for that TNTP palette. 

```{r}
palette_tntp_scales(palette = "likert_5pt")
```

You can use these scale palettes as fill or color aesthetics in ggplot with `scale_fill_tntp` and `scale_color_tntp`.

```{r scale_fill_tntp, fig.width=7, fig.align='center', warning=FALSE}

data.frame(question = "To what extent do you agree...", 
           response = c(rep("Strongly disagree", 3), 
                        rep("Disagree", 4), 
                        rep("Somewhat disagree", 3), 
                        rep("Somewhat agree", 4),  
                        rep("Agree", 10), 
                        rep("Strongly agree", 2))) %>% 
  mutate(response = response %>% factor(levels = rev(c("Strongly disagree", 
                                                   "Disagree", 
                                                   "Somewhat disagree", 
                                                   "Somewhat agree",
                                                   "Agree", 
                                                   "Strongly agree")))) %>% 
  ggplot(aes(question, fill = response)) + 
    geom_bar(position = position_fill()) + 
    theme_tntp_2018(axis_text = "Y", grid = FALSE) + 
    labs(x = NULL, y = NULL, 
         fill = "Response") + 
    coord_flip() + 
    scale_fill_tntp(palette = "likert_6pt")

```
